Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper-results, update results

\Genetics2level{Exp2}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `Reduced' set, transcriptomic data were the best single predictor for
ADL, PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for STA and DMC\@.
For FAT and SUG, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the reduced set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from Exp1 for
  which both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:InbredResults}
\end{figure}

Predictive abilities obtained with the different core sets varied widely within
most traits (Fig.~\ref{fig:InbredResults}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 10\% to 20\%/30\% was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were better predicted via G than via T in
the reduced set of lines, predictive abilities were higher when only a small
number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted better when using T
individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).

\section*{Discussion}
\Genetics2level{Advantages of marker effect models}
Single step approaches in hybrid populations have so far been limited to pig
breeding~\cite{Xiang2015,Xiang2016,Tusell2016}.
All of these studies used breeding value-based single step BLUP methods
(SS-BLUP) conceived by \citeNP{Legarra2009} and \citeNP{Christensen2010} with
extensions to hybrid populations~\cite{Christensen2014,Christensen2015}.
An attactive property of SS-BLUP models is their computational advantage over 
marker-effect single step Bayesian regression models (SSBR) when the number of 
genotypes is smaller than the number of features of the incomplete predictor.
Nevertheless, SS-BLUP suffers from a variety of issues, which we will present
briefly:
First, the combination of the numerator with the genomic relationship matrix
requires commensurability of the two matrices, which necessitates some form of
scaling and weighting of $\mathbf{A}$ and $\mathbf{G}$
\cite{Christensen2012,Christensen2012a}.
Second, the fact that mostly recent individuals have genotypes whereas the 
majority of old individuals has only pedigree records introduces a bias that 
has to be accounted for~\cite{Vitezica2011,Legarra2015,Garcia-Baccino2017}.
Third, the addition of new genotypes requires updates of $\mathbf{G}$ that
demand additional measures such as computing the inverse of $\mathbf{G}$ via
recursion
\cite{Misztal2014,Misztal2016,Misztal2016a,Fragomeni2015,Masuda2016,Pocrnic2016}.
Advantages of the SSBR algorithm, conceived by \citeNP{Fernando2014}, include
that i) it does not require commensurability of $\mathbf{A}$ and $\mathbf{G}$, 
ii) it models the bias incurred by the incomplete predictor explicitly as 
$\mathbf{\epsilon}$ and iii) its speed depends largely on the number of 
features.
To our best knowledge, this study presents the first implementation of the SSBR 
algorithm on a hybrid data set.

>>>>

\Genetics2level{Exp2}
The maize hybrid data set encompasses seven agronomic traits and three
predictors, namely pedigree data (P), genomic information (G), transcriptomic
information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `Reduced' set, transcriptomic data were the best single predictor for
ADL, PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for STA and DMC\@.
For FAT and SUG, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G increased
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the reduced set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from Exp1 for
  which both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:core-results}
\end{figure}

Predictive abilities obtained with the different core sets varied widely within
most traits (Fig.~\ref{fig:core-results}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 10\% to 20\%/30\% was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were better predicted via G than via T in
the reduced set of lines, predictive abilities were higher when only a small
number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted better when using T
individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).

\section*{Discussion}
\Genetics2level{Advantages of marker effect models}
Single step approaches in hybrid populations have so far been limited to pig
breeding~\cite{Xiang2015,Xiang2016,Tusell2016}.
All of these studies used breeding value-based single step BLUP methods
(SS-BLUP) conceived by \citeNP{Legarra2009} and \citeNP{Christensen2010} with
extensions to hybrid populations~\cite{Christensen2014,Christensen2015}.
An attactive property of SS-BLUP models is their computational advantage over 
marker-effect single step Bayesian regression models (SSBR) when the number of 
genotypes is smaller than the number of features of the incomplete predictor.
Nevertheless, SS-BLUP suffers from a variety of issues.
First, the combination of the numerator with the genomic relationship matrix
requires commensurability of the two matrices, which necessitates some form of
scaling and weighting of $\mathbf{A}$ and $\mathbf{G}$
\cite{Christensen2012,Christensen2012a}.
Second, the fact that mostly recent individuals have genotypes whereas the 
majority of old individuals has only pedigree records introduces a bias that 
has to be accounted for~\cite{Vitezica2011,Legarra2015,Garcia-Baccino2017}.
Third, the addition of new genotypes requires updates of $\mathbf{G}$ that
demand additional measures such as computing the inverse of $\mathbf{G}$ via
recursion
\cite{Misztal2014,Misztal2016,Misztal2016a,Fragomeni2015,Masuda2016,Pocrnic2016}.
Advantages of the SSBR algorithm, conceived by \citeNP{Fernando2014}, include
that i) it does not require commensurability of $\mathbf{A}$ and $\mathbf{G}$, 
ii) it models the bias incurred by the incomplete predictor explicitly as 
$\mathbf{\epsilon}$ and iii) its speed depends largely on the number of 
features.
To our best knowledge, this study presents the first implementation of the SSBR 
algorithm on a hybrid data set.
